Sparse Spectral Methods and P-Finite Element Methods on Disk-Splices and Rhombi - Part I: Zero Boundary Conditions

### 2018-12-13 papers/halfdisk/halfdisk.tex
Further edits to FE stuff mainly

\newcommand{\Pnk}{P_{n,k}}
\newcommand{\Pnkab}{P_{n,k}^{(a,b)}}
\newcommand{\Wab}{{W^{(a,b)}}}
\newcommand{\Pmjab}{P_{m,j}^{(a,b)}}
\newcommand{\alphaab}{\alpha^{(a,b)}}
\newcommand{\betaab}{\beta^{(a,b)}}
\newcommand{\bigPab}{\bigP^{(a,b)}}
\newcommand{\Dnt}{D^T_n}
\newcommand{\Wii}{W^{(1,1)}}
\newcommand{\Pii}{P^{(1,1)}}
\newcommand{\Poo}{P^{(0,0)}}
\newcommand{\Poi}{P^{(0,1)}}
\newcommand{\Pio}{P^{(1,0)}}
\newcommand{\dx}{\frac{\partial}{\partial x}}
\newcommand{\dy}{\frac{\partial}{\partial y}}
\newcommand{\element}{\tau}
\newcommand{\refelement}{\hat{\tau}}
\newcommand{\FEset}{\mathcal{T}}

@@

This provides us with a family of orthogonal polynomial spaces on the half disk. Note that we can obtain OP families on other disk-splices and rhombi by appropriately choosing the functions and constants in (2). \bstodo{Should I add in the rhombi / other disk splices formulations here?}

@@

We can develop a quadrature rule on \(\Omega\) in order to evaluate (40) and obtain the coefficients vector \(\mathbf{f}\) as follows. \bstodo{Should I add in the rhombi / other disk splices formulations here?}

@@

We can obtain the matrix operator for \(\Delta\), that will take us from coefficients for expansion in the weighted space
\[
\Wii(x,y) \: \bigP^{(1,1)}(x,y)
\]
to coefficients in the non-weighted space
\[
\bigP^{(1,1)}(x,y).
\]
Importantly, this operator will have Banded-Block-Banded structure, and hence will be sparse. Note that this construction will ensure the imposition of the Dirichlet zero boundary conditions on $\Omega$.

For general \(a,b\) and general function \(f(x,y) = \mathbf{f}^T \: \bigPab(x,y)\), there exist Banded-Block-Banded matrix operators \(D_x^{(a,b)}, \: D_y^{(a,b)}, \: W_x^{(a,b)}, \: W_y^{(a,b)}\) such that:
\begin{align}
\dx[f(x,y)] &= \bigP^{(a+1,b+1)}(x,y)^T \: D_x^{(a,b)} \: \mathbf{f}, \\
\dy[f(x,y)] &= \bigP^{(a,b+1)}(x,y)^T \: D_y^{(a,b)} \: \mathbf{f}, \\
\dx[\Wab(x,y) \: f(x,y)] &= W^{(a-1,b-1)}(x,y) \: \bigP^{(a-1,b-1)}(x,y)^T \: W_x^{(a,b)} \: \mathbf{f}, \\
\dy[\Wab(x,y) \: f(x,y)] &= W^{(a,b-1)}(x,y) \: \bigP^{(a,b-1)}(x,y)^T \: W_y^{(a,b)} \: \mathbf{f}.
\end{align}
The incrementing and decrementing of parameters as seen here is analogous to other well known orthogonal polynomials families' derivatives, for example the Jacobi polynomials, as seen in the DLMF \cite{DLMFDerivatives}.
Further, we note that there exist Banded-Block-Banded matrix operators \(T^{(1,0)\to(0,0)}\) and \(T^{(0,1)\to(1,1)}\) such that:
\begin{align}
x \bigP^{(1,0)}(x,y) &= \Big(T^{(1,0)\to(0,0)} \Big)^T \: \bigP^{(0,0)}(x,y) \\
\bigP^{(0,1)}(x,y) &= \Big(T^{(0,1)\to(1,1)} \Big)^T \: \bigP^{(1,1)}(x,y)
\end{align}
Then, the operator \(\Delta\) is given by
\begin{align}
    \Delta := D_x^{(0,0)} \: W_x^{(1,1)} + T^{(0,1)\to(1,1)} \: D_y^{(0,0)} \: T^{(1,0)\to(0,0)} \: W_y^{(1,1)}.
\end{align}
The sparsity of the differential operators can be shown using a simply integration by parts argument, again showing that the sparsity for differential operators is not dependent on the domain.

%
\section{P-finite element methods using sparse operators}
\bstodoinline{Need to discuss this}
We follow the method of \cite{beuchler2006new}. Consider the 2D Dirichlet problem on a domain $\Omega$:
\begin{align}
	\begin{cases}
         - \Delta u(x,y) = f(x,y) \quad \text{in } \Omega \\
         u = 0 \quad \text{on } \partial \Omega
         \end{cases}
\end{align}
This has the weak formulation for any test function $v \in V := H_0^1(\Omega) = \{v \in H^1(\Omega) \quad | \quad v|_{\partial \Omega} = 0 \}$,
\begin{align}
	L(v) := \int_\Omega f \: v \: d\mathbf{x} = \int_\Omega \nabla u \cdot \nabla v \: d\mathbf{x} =: a(u,v).
\end{align}

Let $\FEset$ be the set of elements $\element$ that make up our finite element discretisation of the domain, where each $\element$ is a rhombus for example. We can choose our finite dimensional space $V_h = \{v_h \in V \quad | \quad deg(v_h)|_\element \le p \: \forall \element \in \FEset\}$.

@@

We can choose a basis for $V_h$ by using the orthogonal polynomials 
\begin{align}
\bigP(x,y) := \begin{bmatrix}
		\bigP_0(x,y) \\
		\hline
		\bigP_1(x,y) \\
		\hline
		\bigP_2(x,y) \\
		\hline
		\vdots \\
		\hline
		\bigP_p(x,y)
	\end{bmatrix},
\quad
\bigP_n(x,y) := \begin{bmatrix}
		P_{n,0}(x,y) \\
		\vdots \\
		P_{n,n}(x,y)
	\end{bmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,p,
\end{align}
Appropriately relabelling these OPs as the set $\{\phi_1,\dots,\phi_N\}$, we can define the global stiffness matrix $A$ by $A_{ij} =  a(\phi_i, \phi_j)$. The global stiffness matrix can be computed using the element stiffness matrices $A^\element = \int_\element \nabla \bigP \cdot \nabla \bigP \: d\mathbf{x}, \: \element \in \FEset$. Then we have the linear system for our finite element problem:
\begin{align}
A\mathbf{u} = \mathbf{f}
\end{align}
where the vectors $\mathbf{u}, \mathbf{f}$ are the coefficient vectors for expansion of the solution $u_h$ and for the function $f$ respectively.

Using the quadrature rule we have formulated for integrals on a rhombus element $\element$, and the sparse operator matrices for divergence (partial derivatives), we obtain a sparse global stiffness matrix.

>>>>

\newcommand{\Pnk}{P_{n,k}}
\newcommand{\Pnkab}{P_{n,k}^{(a,b)}}
\newcommand{\Wab}{{W^{(a,b)}}}
\newcommand{\Pmjab}{P_{m,j}^{(a,b)}}
\newcommand{\alphaab}{\alpha^{(a,b)}}
\newcommand{\betaab}{\beta^{(a,b)}}
\newcommand{\bigPab}{\bigP^{(a,b)}}
\newcommand{\Dnt}{D^T_n}
\newcommand{\Wii}{W^{(1,1)}}
\newcommand{\Pii}{P^{(1,1)}}
\newcommand{\bigPii}{{\mathbb{P}^{(1,1)}}}
\newcommand{\Poo}{P^{(0,0)}}
\newcommand{\bigPoo}{{\mathbb{P}^{(0,0)}}}
\newcommand{\Poi}{P^{(0,1)}}
\newcommand{\Pio}{P^{(1,0)}}
\newcommand{\dx}{\frac{\partial}{\partial x}}
\newcommand{\dy}{\frac{\partial}{\partial y}}
\newcommand{\element}{\tau}
\newcommand{\refelement}{\hat{\tau}}
\newcommand{\FEset}{\mathcal{T}}
\newcommand{\bigW}{\mathbb{W}}
\newcommand{\bigWii}{{\mathbb{W}^{(1,1)}}}

@@

This provides us with a family of orthogonal polynomial spaces on the half disk. Note that we can obtain OP families on other disk-splices and rhombi by appropriately choosing the functions and constants in (2).

@@

We can develop a quadrature rule on the half disk \(\Omega\) in order to evaluate (40) and obtain the coefficients vector \(\mathbf{f}\) as follows.

@@

For general \(a,b\) and general function \(f(x,y) = \mathbf{f}^T \: \bigPab(x,y)\), there exist Banded-Block-Banded matrix operators \(D_x^{(a,b)}, \: D_y^{(a,b)}, \: W_x^{(a,b)}, \: W_y^{(a,b)}\) such that:
\begin{align}
\dx[f(x,y)] &= \bigP^{(a+1,b+1)}(x,y)^T \: D_x^{(a,b)} \: \mathbf{f}, \\
\dy[f(x,y)] &= \bigP^{(a,b+1)}(x,y)^T \: D_y^{(a,b)} \: \mathbf{f}, \\
\dx[\Wab(x,y) \: f(x,y)] &= W^{(a-1,b-1)}(x,y) \: \bigP^{(a-1,b-1)}(x,y)^T \: W_x^{(a,b)} \: \mathbf{f}, \\
\dy[\Wab(x,y) \: f(x,y)] &= W^{(a,b-1)}(x,y) \: \bigP^{(a,b-1)}(x,y)^T \: W_y^{(a,b)} \: \mathbf{f}.
\end{align}
The incrementing and decrementing of parameters as seen here is analogous to other well known orthogonal polynomials families' derivatives, for example the Jacobi polynomials, as seen in the DLMF \cite{DLMFDerivatives}.
Further, we note that there exist Banded-Block-Banded matrix operators \(T^{(1,0)\to(0,0)}\) and \(T^{(0,1)\to(1,1)}\) such that:
\begin{align}
x \bigP^{(1,0)}(x,y) &= \Big(T^{(1,0)\to(0,0)} \Big)^T \: \bigP^{(0,0)}(x,y) \\
\bigP^{(0,1)}(x,y) &= \Big(T^{(0,1)\to(1,1)} \Big)^T \: \bigP^{(1,1)}(x,y)
\end{align}

We can then obtain the matrix operator for \(\Delta\), that will take us from coefficients for expansion in the weighted space
\[
\Wii(x,y) \: \bigPii(x,y)
\]
to coefficients in the non-weighted space
\[
\bigPii(x,y).
\]
Importantly, this operator will have Banded-Block-Banded structure, and hence will be sparse. Note that this construction will ensure the imposition of the Dirichlet zero boundary conditions on $\Omega$. The matrix operator \(\Delta\) acting on the coefficients vector is then given by
\begin{align}
    \Delta := D_x^{(0,0)} \: W_x^{(1,1)} + T^{(0,1)\to(1,1)} \: D_y^{(0,0)} \: T^{(1,0)\to(0,0)} \: W_y^{(1,1)}.
\end{align}
The sparsity of the these operators can be shown using a simply integration by parts argument, again showing that the sparsity for differential operators is not dependent on the domain.

%
\section{P-finite element methods using sparse operators}

We follow the method of \cite{beuchler2006new}. Consider the 2D Dirichlet problem on a domain $\Omega$:
\begin{align}
	\begin{cases}
         - \Delta u(x,y) = f(x,y) \quad \text{in } \Omega \\
         u = 0 \quad \text{on } \partial \Omega
         \end{cases}
\end{align}
This has the weak formulation for any test function $v \in V := H_0^1(\Omega) = \{v \in H^1(\Omega) \quad | \quad v|_{\partial \Omega} = 0 \}$,
\begin{align}
	L(v) := \int_\Omega f \: v \: d\mathbf{x} = \int_\Omega \nabla u \cdot \nabla v \: d\mathbf{x} =: a(u,v).
\end{align}

In general, we would let $\FEset$ be the set of elements $\element$ that make up our finite element discretisation of the domain, where each $\element$ is a rhombus or disk slice for example. For the purposes of this paper, we limit this to a single element, that is we let $\element = \Omega$ for a half disk/disk slice/rhombus domain. We can choose our finite dimensional space $V_h = \{v_h \in V \quad | \quad deg(v_h|_\element) \le p\}$ for some $p \in \N$.

@@

Define $\Lambda^{(a,b)} := \langle \bigP^{(a,b)}, \: {\bigP^{(a,b)}}^T \rangle_{W^{(a,b)}}$ where $W^{(a,b)}$ is the weight with which the OPs in $\bigP^{(a,b)}$ are orthogonal with respect to. Note that due to orthogonality this is a diagonal matrix. We can choose a basis for $V_h$ by using the weighted orthogonal polynomials on $\element$ with parameters $a = b = 1$:
\begin{align}
\bigWii(x,y) &:= \begin{bmatrix}
		\bigWii_0(x,y) \\
		\hline
		\bigWii_1(x,y) \\
		\hline
		\bigWii_2(x,y) \\
		\hline
		\vdots \\
		\hline
		\bigWii_p(x,y)
	\end{bmatrix}, \\
\bigWii_n(x,y) &:= \begin{bmatrix}
		\Wii(x,y) \: \Pii_{n,0}(x,y) \\
		\vdots \\
		\Wii(a,y) \: \Pii_{n,n}(x,y)
	\end{bmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,p,
\end{align}
and rewrite (63) in matrix form:
\begin{align}
	a(u_h,v_h) &= \int_\element \nabla u_h \cdot \nabla v_h \: d\mathbf{x} \\
	&= \int_\element \begin{bmatrix}
					\partial_x v_h \\
					\partial_y v_h
				\end{bmatrix}^T 
				\begin{bmatrix}
					\partial_x u_h \\
					\partial_y u_h
				\end{bmatrix}
				\\
	&= \int_\element \begin{bmatrix}
					\bigPoo^T \Wii_x \mathbf{v} \\
					\bigPoo^T T^{(1,0)\to(0,0)} \Wii_y \mathbf{v}
				\end{bmatrix}^T 
				\begin{bmatrix}
					\bigPoo^T \Wii_x \mathbf{u} \\
					\bigPoo^T T^{(1,0)\to(0,0)} \Wii_y \mathbf{u}
				\end{bmatrix}
				\\
	&= \int_\element \Big( \mathbf{v}^T {\Wii_x}^T \bigPoo \bigPoo^T \Wii_x \mathbf{u} \nonumber \\
					& \quad \quad \quad + \mathbf{v}^T ({T^{(1,0)\to(0,0)} \Wii_y})^T \bigPoo \bigPoo^T T^{(1,0)\to(0,0)} \Wii_y \mathbf{u}  \Big) \: d\mathbf{x} \\
	&= \mathbf{v}^T \: \Big( {\Wii_x}^T \Lambda^{(0,0)} \Wii_x + ({T^{(1,0)\to(0,0)} \Wii_y})^T \Lambda^{(0,0)} T^{(1,0)\to(0,0)} \Wii_y \Big) \: \mathbf{u}
\end{align}
where $\mathbf{u}, \mathbf{v}$ are the coefficient vectors of the expansions of $u_h, v_h \in V_h$ respectively in the $V_h$ basis ($\bigWii$ OPs), and
\begin{align}
	L(v_h) &= \int_\element \: v_h \: f \: d\mathbf{x} \\
	&= \int_\element \: \mathbf{v}^T \: \bigWii \: \bigPii^T \: \mathbf{f} \: d\mathbf{x} \\
	&= \mathbf{v}^T \Lambda^{(1,1)} \: \mathbf{f},
\end{align}
where $\mathbf{f}$ is the coefficient vector for the expansion of the function $f(x,y)$ in the $\bigPii$ OP basis. We can also not that naturally $\langle \bigPii, {\bigPii}^T \rangle_{\Wii}$ is just a diagonal matrix.

Since (63) is equivalent to stating that
\begin{align}
	L(\Wii \Pii_{n,k}) = a(u_h,\Wii \Pii_{n,k}) \quad \forall \: n = 0,\dots,p, \: k = 0,\dots,n,
\end{align}
(i.e. holds for all basis functions of $V_h$) by choosing $v_h$ as each basis function, we can equivalently write the linear system for our finite element problem as:
\begin{align}
A\mathbf{u} = \tilde{\mathbf{f}}.
\end{align}
where the (element) stiffness matrix $A$ is defined by 
\begin{align}
A = {\Wii_x}^T \Lambda^{(0,0)} \Wii_x + ({T^{(1,0)\to(0,0)} \Wii_y})^T \Lambda^{(0,0)} T^{(1,0)\to(0,0)} \Wii_y, 
\end{align}
and the load vector $\tilde{\mathbf{f}}$ is given by 
\begin{align}
\tilde{\mathbf{f}} = \Lambda^{(1,1)} \: \mathbf{f}.
\end{align}

Note the since we have sparse operator matrices for partial derivatives and basis-transform, we obtain a symmetric sparse (element) stiffness matrix, as well as a sparse operator matrix for calculating the load vector (rhs).
